A Spectral Analysis of the Domain Decomposed Monte Carlo Method for Linear Systems

### 2015-04-13 doc/ned2015/response_to_reviewer.txt

Response to reviewer's comments:

------------------------------------------------------------------------------
1. Since the underlying analysis seems to be graphical, they should use as
many graphs to explain what is happening.  Following words and moving
information is not so easy for a reader.

We agree that without visual aids the discussion is complex and is very much
based on the underlying graph of the matrix. We have added a figure (Figure 1)
in section 3.1 and discussion the random walk in terms of the sparse graph
underneath the matrices that should help clarify the explanation, particularly
for the n_s and n_i parameters in our diffusion analysis. In addition, we have
expanded our description of the algorithm in section 2 and have provided more
definitions for clarity.

------------------------------------------------------------------------------
2. Abstract: The term "measured computational results" is not so clear at
first.  The authors may want to find another name for this quantity.

We agree and have clarified that leakage fractions computed via numerical
experiments show good agreement with the equations we have derived.

------------------------------------------------------------------------------
3. Page 2: Last sentence of the second paragraph, it is not clear what is
meant by "specifically, have poor algorithmic strong scaling."

This statement has been clarified to indicate that when an algorithm is strong
scaled, the amount of on-process work decreases while the amount of
communication increases due to increasing processor count and decreasing local
problem size.

------------------------------------------------------------------------------
4. Top of Page 3: "adjoint method Monte Carlo method", is it necessary to use
'method' twice?  Seems a bit awkward.

This is a typo and it has been corrected.

------------------------------------------------------------------------------
5. Should indicate after Eq(2) .. yet to be determined adjoint source…

We agree and have indicated that these quantities will be determined in the
derivation.

------------------------------------------------------------------------------
4.  I assume Wm is a weight vector.

Wm is a scalar quantity equal to the weight of a random walk on its m^th
step. We have clarified the definition of Wm by indicating this.

------------------------------------------------------------------------------
6. D is the diffusion FD operator.

We have clarified that D is the diffusion operator resulting from our choice
of finite difference scheme.

------------------------------------------------------------------------------
7. Why not diagonalize D when it first appears?

We agree that we could diagonalize D initially. However, for more difficult
problems diagonal preconditioning may not be sufficient to satisfy the
convergence criteria and therefore it is instructive to separate it in the
analysis to show how a preconditioner would affect both iterative performance
as well as indicate how our analysis could be extended for more complicated
preconditioners. In addition, for our model problem the preconditioner simply
results in a scaling of the operator spectrum computed in section 4.1 and
therefore lets us analyze the operator independently from the preconditioner.

------------------------------------------------------------------------------
8. I believe D can be written as a block tridiagonal matrix for the stencil
used.  Is this the case?  If not see Burlish and Stoerer.  This would simplify
the analysis greatly.

Although we agree other matrix orderings could have been used, the results of
the paper are fundamentally based on the local stencil and independent the
overall global structure of the matrix. A block tridiagonal ordering could
reduce the bandwidth of the matrix and improve computational efficiency for a
matrix-vector multiply. However, the Monte Carlo algorithm would perform
identically in both cases given the same parallel decomposition of the
problem. The addition of Figure 1 in response to the first comment should also
help with this part of the analysis.

For example, given a banded ordering like that used in this work or the
suggested block tridiagonal ordering, the probability distribution functions
would be of the same size and have the same nearest-neighbor connections
because both use the same local stencil. Therefore, both cases will yield
random walks of the same time and space complexity. In addition, if both cases
had the same domain decomposition, random walks would be communicated from
domain to domain in an identical matter because the same local stencil was
used, therefore also resulting in the same parallel analysis.

------------------------------------------------------------------------------
9. Finally, there are no flux results with estimated errors.  We need results
since, after all, they are the bottom-line.

The purpose of the paper is to provide analytic relationships for the
domain-to-domain communication of random walks, and therefore a measure of
potential bandwidth and latency issues, when the Neumann-Ulam algorithm is
subject to domain decomposition. The domain decomposition itself does not
perturb the results of the stochastic algorithm (e.g. the flux in this
case). Rather, it simply provides a means to perform random walks via the
rules of the Neumann-Ulam game where the data for the random walks (weights
and probablities) has been domain-decomposed.

Variance estimation, however, is a different story and domain decomposition
schemes to date trade inexact variance estimates for scalablity. This is due
to the fact that the order of events must be maintained to acheive a correct
variance estimate and fully asynchronous algorithms do not, in general, permit
this. The literature indicates that this order of operations is less
significant for the correctness of the flux estimate with differences in
computations induced by round-off error in floating point
computations. Furthermore, the accuracy and convergence properties of the
Monte Carlo method have been well documented in the references cited by the
paper.

Therefore, because the purpose of this work is to analyze communication costs
of the algorithm using an operator with spectral properties similar to large
scale neutronics problems and not computing the global flux and variance
estimate, we do not feel it is necessary in this paper to report flux results.

We will, however, in an upcoming paper present our full domain decomposition
algorithm for parallelizing the solver and give a thorough analysis of its
correctness and scalability.

------------------------------------------------------------------------------

### 2015-04-13 doc/ned2015/slattery_wilson_evans.tex
updating ned paper for reviewer comments

%%---------------------------------------------------------------------------%%
  \begin{abstract}
    The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
    method for solving linear systems is analyzed using the spectral
    properties of the linear operator. Relationships for the average length of
    the adjoint random walks, a measure of convergence speed and serial
    performance, are made with respect to the eigenvalues of the linear
    operator. In addition, relationships for the effective optical thickness
    of a domain in the decomposition are presented based on the spectral
    analysis and diffusion theory. Using the effective optical thickness, the
    Wigner rational approximation and the mean chord approximation are applied
    to estimate the leakage fraction of random walks from a domain in the
    decomposition as a measure of parallel performance and potential
    communication costs. The one-speed, two-dimensional neutron diffusion
    equation is used as a model problem in numerical experiments to test the
    models for symmetric operators with spectral qualities similar to light
    water reactor problems. In general, the derived approximations show good
    agreement with effective optical thicknesses and leakage fractions
    computed by numerical experiments.
\end{abstract}

@@

An alternative approach to approximate matrix inversion is to employ Monte
Carlo methods that sample a distribution with an expectation value equivalent
to that of the inverted operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that, for certain classes of problems, are competitive with
conventional methods methods and are applicable to a wide variety of linear
systems \cite{evans_monte_2009,evans_monte_2014}. To advance the applicability
of these new MCSA methods, we therefore aim to apply the underlying Monte
Carlo methods to large-scale problems in parallel computing environments.

@@

We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:.
  \label{eq:linear_problem}
\end{equation}
Choosing the adjoint Monte Carlo method to invert the linear operator, we
define the linear system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the yet to be determined adjoint solution and
source respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
Monte Carlo estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We first rearrange Eq~(\ref{eq:adjoint_linear_problem}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ is the \textit{iteration matrix}.
This yields the adjoint Neumann series:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
eigenvalues and therefore has the same spectral radius. We expand this
summation to yield a series of transitions that can be approximated by
a Monte Carlo random walk sequence:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
To find the solution to Eq~(\ref{eq:linear_problem}) we build an
estimator for the adjoint solution from this series expansion. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

In order to perform the random walks required to approximate the
summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build the
column-scaled transition probability matrix and associated transition
weights as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{|h_{ji}|}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
such that for each step, $m$, in the random walk, we expect to select a new
state, $j$, from the current state, $i_m$, by sampling the $i_m^{th}$ row of
the probability matrix as a discrete distribution, and updating the weight of
the random walk at the $m^{th}$ step, $W_m$, accordingly.  Using our result
from Eq~(\ref{eq:inner_product_constraint}) generated by applying the adjoint
constraints, the Monte Carlo estimator for the solution in state $i$ for a
particular random walk permutation, $\nu$, is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in the
current state and $W_m$ is the random walk weight on the $m^{th}$
step. Finally, the expectation value using all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{m=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_m}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{m-1},i_m} \delta_{i_m,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
where $b_{i_0}$ will be the sampled source starting weight, therefore
also providing an unbiased Monte Carlo estimate of the solution.

We also desire a criteria for random walk termination for problems where only
an approximate solution is necessary. We utilize a relative weight cutoff
parameter:
\begin{equation}
  W_f = W_c W_0:\,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff} and $W_0$ is the initial weight of
the random walk. The adjoint random walk will then be terminated after $m$
steps if $W_m < W_f$ as tally contributions become increasingly small,
effectively establishing a measure for the degree of approximation of the
solution.

@@

We choose a finite difference scheme on a square Cartesian grid to discretize
the problem. We choose the 9-point stencil over a grid of size $h$
\cite{leveque_finite_2007} for the Laplacian:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator resulting from the finite difference
discretization, $\ve{s}$ is the source in vector form and $\boldsymbol{\phi}$
is the vector of unknown fluxes.

%%---------------------------------------------------------------------------%%
\subsection{Domain Decomposition of the Model Problem}
\label{subsec:model_problem_decomp}
To consider domain-to-domain communication in the parallel Monte Carlo
algorithm, we must consider the required data with the correct parallel
decomposition. In the adjoint Neumann-Ulam method, the domain decomposition
consists of all states in the system that are local, and the probabilities and
weights for all state transitions possible in the local domain. Given an input
parallel decomposition for the linear operator, by definition in
Eq~(\ref{eq:adjoint_probability}) probabilities and weights will have the same
parallel decomposition and underlying sparse graph as the input matrix.

@@

We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi preconditioning
is sufficient to reduce the spectral radius of the iteration matrix below
unity and therefore ensure convergence of the Neumann series. The
preconditioner in this case is then $\ve{M} = diag(\ve{D})$ such that we are
solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion operator
with each row scaled by the diagonal component. As we have defined a
homogeneous domain, the scaling factor, $\alpha$, is the same for all rows in
the operator and defined as the $\phi_{i,j}$ coefficient from
Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

>>>>

  %%---------------------------------------------------------------------------%%
  \begin{abstract}
    The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
    method for solving linear systems is analyzed using the spectral
    properties of the linear operator. Relationships for the average length of
    the adjoint random walks, a measure of convergence speed and serial
    performance, are made with respect to the eigenvalues of the linear
    operator. In addition, relationships for the effective optical thickness
    of a domain in the decomposition are presented based on the spectral
    analysis and diffusion theory. Using the effective optical thickness, the
    Wigner rational approximation and the mean chord approximation are applied
    to estimate the leakage fraction of random walks from a domain in the
    decomposition as a measure of parallel performance and potential
    communication costs. The one-speed, two-dimensional neutron diffusion
    equation is used as a model problem in numerical experiments to test the
    models for symmetric operators with spectral qualities similar to light
    water reactor problems. In general, the derived approximations show good
    agreement with random walk lengths and leakage fractions computed by the
    numerical experiments.
  \end{abstract}

@@

An alternative approach to solving linear systems is to employ Monte Carlo
methods that sample a distribution with an expectation value equivalent to
that of the inverted linear operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that, for certain classes of problems, are competitive with
conventional methods methods and are applicable to a wide variety of linear
systems \cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA
methods to engineering problems, our research aims to improve the underlying
Monte Carlo algorithms to handle large-scale problems in parallel computing
environments.

@@

We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:.
  \label{eq:linear_problem}
\end{equation}
Choosing the adjoint Monte Carlo method to invert the linear operator, we
define the linear system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the yet to be determined adjoint solution and
source respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
Monte Carlo estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We first rearrange Eq~(\ref{eq:adjoint_linear_problem}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ is the \textit{iteration matrix}.
This yields the adjoint Neumann series:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of $\ve{H}$
must remain less than unity\footnote{$\ve{H}^T$ contains the same eigenvalues
  as $\ve{H}$ and therefore has the same spectral radius.}. We expand the
matrix-vector multiply element-wise in this summation to yield a sum of sums:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
To find the solution to Eq~(\ref{eq:linear_problem}) we build an
estimator for the adjoint solution from this series expansion. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we require two
constraints to close the system. We use Eq~(\ref{eq:adjoint_vector_relation})
as the first constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the $j^{th}$
component is the Kronecker delta function, $\delta_{i,j}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint in
Eq~(\ref{eq:adjoint_vector_relation}) we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source and the
adjoint solution using a delta function source, we recover one component of
the original solution.

In the Neumann-Ulam method, a Monte Carlo game is played to construct the
terms of the Neumann series in Eq~(\ref{eq:adjoint_neumann_solution}). We use
random walks of length $k$ to build the series of state transitions in the
summation:
\begin{equation}
  \nu = i_0 \rightarrow i_1 \rightarrow i_2 \rightarrow \cdots \rightarrow
  i_{k-1}\:,
  \label{eq:permutation}
\end{equation}
where $\nu$ is a particular random walk permutation. Starting a random walk
permutation requires the computation of an initial starting state, $i_0$, and
initial weight, $W_0$. The initial state, $i_0$, of the random walk is
determined by sampling the source vector $\ve{b}$ with probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
and with a corresponding starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_{i_0}}{|b_{i_0}|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
After a random walk is started, we construct the remaining terms in
Eq~(\ref{eq:adjoint_neumann_solution}) using a row-scaled transition probability
matrix and corresponding transition weights:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{|h_{ji}|}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
where each row of the probability matrix, $\ve{P}_i$, gives a probability
distribution function for moving from row $i$ to a new row, $j$. As a random
walk is computed by sampling the probability matrix, every stochastic
transition from a state $i$ to a state $j$ has a corresponding weight,
$w_{ij}$, that is used to update the current weight of the random walk. At the
$m^{th}$ step of a random walk, its weight is:
\begin{equation}
  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
with $W_0$ given by Eq~(\ref{eq:adjoint_starting_weight}). To stop a random
walk, we utilize a relative weight cutoff parameter:
\begin{equation}
  W_f = W_c W_0:\,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff}. The random walk will be terminated
after $m$ steps if $W_m < W_f$.

Using our result from Eq~(\ref{eq:inner_product_constraint}) the Monte Carlo
estimator for the solution in state $i$ for a particular random walk
permutation of length $k$ is:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in the
current state, $i_m$. Finally, the expectation value using all random walk
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded with Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_source_definition}) directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:.
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}

@@

We choose a finite difference scheme on a square Cartesian grid to discretize
the problem. We choose the 9-point stencil over a grid of size $h$ for the
Laplacian \cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator resulting from the finite difference
discretization, $\ve{s}$ is the source in vector form and $\boldsymbol{\phi}$
is the vector of unknown fluxes.

%%---------------------------------------------------------------------------%%
\subsection{Domain Decomposition of the Model Problem}
\label{subsec:model_problem_decomp}
To investigate domain-to-domain communication in a parallel Monte Carlo
algorithm, we must consider the required data with the correct parallel
decomposition. In the adjoint Neumann-Ulam method, the domain decomposition
consists of all states in the system that are local, and the probabilities and
weights for all state transitions possible in the local domain. Given an input
parallel decomposition for the linear operator, by definition in
Eq~(\ref{eq:adjoint_probability}) the probabilities and weights will have the
same parallel decomposition and underlying sparse graph as the input
matrix. In the case of the model problem, parallel decomposition of states and
subsequently the input matrix is determined by a spatial partitioning of the
grid.

@@

We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$
on a rectilinear Cartesian grid:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi preconditioning
is sufficient to reduce the spectral radius of the iteration matrix below
unity and therefore ensure convergence of the Neumann series. The
preconditioner in this case is $\ve{M} = diag(\ve{D})$ such that we are
solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion operator
with each row scaled by the diagonal component. As we have defined a
homogeneous domain, the scaling factor, $\alpha$, is the same for all rows in
the operator and defined as the $\phi_{i,j}$ coefficient from
Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we have the following spectrum of preconditioned
eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

The spectral radius of the iteration matrix is obtained by seeking its largest
eigenvalue. As with the diffusion operator, we can use the same analysis
techniques to find the eigenvalues for the iteration matrix. We use a few
simplifications by noting that if the Jacobi preconditioned iteration matrix
is $\ve{H} = \ve{I} - \ve{M}^{-1}\ve{D}$, then we expect all terms on the
diagonal of the iteration matrix to be zero such that we have the following
stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the eigenfunctions defined by Eq~(\ref{eq:eigenfunction_form}) we
get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum of the Jacobi preconditioned iteration
matrix. We find that the maximum eigenvalue exists when $p=q=0$ giving the
following spectral radius of the Jacobi preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}
